Alone Together: Compositional Reasoning and Inference for Weak Isolation

### 2017-07-08 inference.tex
Edits to Sec. 2.
Edits
changes to R_l
Edits. URL added.

The Rely-Guarantee framework presented in the previous section
facilitates modular proofs for weakly-isolated transactions, but
imposes a non-trivial annotation burden.  In particular, it requires
each statement ($c$) of the transaction to be annotated with a stable
pre- ($P$) and post-condition ($Q$), and loops to be annotated with
stable inductive invariants ($\psi$). While weakest pre-condition
style predicate transformers can help in inferring intermediate
assertions for regular statements, loop invariant inference remains
challenging, even for the simple form of loops considered here.  As an
alternative, we present an inference algorithm based on state
transformers.  The idea is to infer the logical effect that each
statement has on the transaction-local database state $\stl$ (i.e.,
how it transforms $\stl$), and compose multiple such effects together
to describe the effect of the transaction as a whole.  Importantly,
this approach generalizes to loops, where the effect of a loop can be
inferred as a well-defined function of the effect of its body, thanks
to certain pleasant properties enjoyed by the database programs
modeled by $\txnimp$.  Interpreting database semantics as functional
transformations on sets (described in terms of their logical effects)
enables an inference mechanism that can leverage off-the-shelf SMT
solvers for automated verification.

@@

At the core of our approach is a simple language ($\SL$) to express
set transformations (see Fig.~\ref{fig:logic-syntax}). The language
admits only set expressions that include variables ($x$), literals of
the form $\{x \,|\, \varphi\}$ where $\varphi$ is a propositional
(quantifier-free) formula on $x$, a restricted form of existential
quantification that binds a set $x$ satisfying proposition $\phi$ in a
set expression $s$, a monadic composition of two set expressions
($s_1$ and $s_2$) composed using a bind ($\bind$) operation, a
conditional set expression where the condition is a propositional
formula, and a union of two set expressions. Symbols $\stl$ and $\stg$
are also variables in $\SL$, but are used to denote local and database
states, respectively.  The language is carefully chosen to be
expressive enough to capture the semantics of $\txnimp$ statements (as
well as SQL operations more generally), yet simple enough to have a
semantics-preserving translation to a decidable fragment of
first-order logic.

@@

where, $\bar{v}$ are the variables that occur free in $\F$; this is

@@

equivalence in $\SL$ translates to the equality in first-order logic,

@@

positive ($I(\stg) \Leftrightarrow \forall(r\in\stg).~r.\C{bal}\ge

@@

The state transformer rules, like the earlier RG rules, closely follow
the corresponding reduction rules in Fig.~\ref{fig:txnimp}, except
that their language of expression is $\SL$. For instance, while the
reduction rule for \C{UPDATE} declaratively specifies the set of
updated records, the state transformer rule uses $\SL$'s bind
operation to \emph{compute} the set. Other SQL rules do likewise. The
rules for \C{LET} binders, conditionals, and sequences compose the
effects inferred for their subcommands. For instance, the effect of a
sequence of commands $c_1;c_2$ is the union of effects $F_1$ and $F_2$
of $c_1$ and $c_2$, respectively, except that $F_2$ is applied to a
state ($\stl$) to which $F_1$ has already been applied, reflecting
their order of reduction. The inference rule for \C{FOREACH} takes
advantage of the $\SL$'s bind operator to lift the effect inferred for
the loop body to the level of the loop. Since records added to $\stl$
in each iteration of \C{FOREACH} are independent of the previous
iteration, sequential composition of the effects of different
iterations is same as their parallel composition. Since the loop body
is executed once per each $z\in x$, the effect of the the loop is the
union of effects ($\F$) for all $z\in x$, all applied to the same
state ($\stl$ and $\stg$). That is, $\F_{loop}(\stl,\stg) =
\bigcup_{z\in x}\F_{body}(\stl,\stg)$. From the definition of the set
monad's bind operator, $\F_{loop}(\stl,\stg) = x \bind (\lambda
z.~F_{body}(\stl,\stg))$, which mirrors the definition of the rule.

@@

Theorem~\ref{thm:inference-sound} lets us replace the local judgment
of the \rulelabel{RG-Txn} rule (Fig.~\ref{fig:rg-rules}) by a state
transformer inference judgment. The soundness of a transaction's guarantee can now
be established w.r.t the effect $\F$ of the body. The
\rulelabel{RG-Txn} rule so updated is shown
below:
\begin{smathpar}
\begin{array}{c}
\RULE
{
  \stable(R,\I)\spc
  \stable(R,I)\\
  \R(\stl,\stg,\stg') \Leftrightarrow \I(\C{true},\stl,\stg,\stg') 
        \wedge R(\stg,\stg') \spc
  c \elabsto \F\\
  \R_c(\stl,\stg,\stg') \Leftrightarrow \I(\C{false},\stl,\stg,\stg') 
        \wedge R(\stg,\stg') \spc
  \stable(\R_c,\F)\\
  \forall \stg.~ G(\stg, F(\emptyset,\stg) \gg \stg)\spc
  \forall \stg,\stg'.~I(\stg) \wedge G(\stg,\stg') \Rightarrow I(\stg')\\
}
{
  \rg{I,R}{\ctxn{i}{\I}{c}}{G,I}
}
\end{array}
\end{smathpar}
Automating the application of the \rulelabel{RG-Txn} rule for a
transaction requires automating the multiple implication checks in
the premise. While $R$, $\R$, $\R_c$, $\I$ and $I$ are formulas in
first-order logic (FOL) with a relatively simple structure, $\F$
is an expression in the set language $\SL$
(Fig.~\ref{fig:logic-syntax}) with a possibly complex structure.
Fortunately, however, there exists a semantics-preserving translation
from $\SL$ to a restricted subset of first-order logic (FOL) that
lends itself to automatic reasoning.

@@

Observe that the encoding shown in Fig.~\ref{fig:logic} maps to
a subset of logic that satisfies the following syntactic properties:
\begin{itemize}
  \item All quantification is first-order; second-order objects, such
    as sets and functions, are never quantified.
  \item Quantifiers appear only at the prenex position, i.e., at the
    beginning of a quantified formula.  
  \item All function symbols, modulo those that might appear in
    $\varphi\in\Prop^0$, are uninterpreted and Boolean.
\end{itemize}
This fragment of FOL known as EPR (Effectively Propositional logic)
is known to be decidable~\cite{z3epr}. The language of encoding, however, is a
combination of {\sf EPR} with (a). $\Prop^0$, the theory from which
quantifier-free propositions ($\varphi$) that encode object language
expressions are drawn, and (b).  $\Prop^1$, the theory from which
invariants ($I$) are drawn. We write $\SL[\Prop^0,\Prop^1]$ to
highlight the parameterization of $\SL$ on $\Prop^0$ and $\Prop^1$,
and state the following theorem:
\begin{theorem}
  $\SL[\Prop^0,\Prop^1]$ is decidable if ${\sf EPR}+\Prop^0+\Prop^1$
  is decidable.
\end{theorem}
A useful instantiation of $\SL$ is $\SL[{\sf SLA},{\sf EPR}+{\sf
SLA}]$, where ${\sf SLA}$ is the theory of simple linear arithmetic.
Since {\sf EPR}+{\sf SLA} is known to be decidable~\cite{eprsla}, hence:
\begin{theorem}
  $\SL[{\sf SLA},{\sf EPR}+{\sf SLA}]$ is decidable.
\end{theorem}
The $\SL[{\sf SLA},{\sf EPR}+{\sf SLA}]$ instantiation requires $I$ to
be drawn from {\sf EPR}+{\sf SLA}, which is expressive enough to
describe common database integrity constraints, such as referential
integrity, non-negativeness of all integer values in a column etc.
The isolation specifications presented in Sec.~\ref{sec:isolation} are
already simple first-order formulas that can be encoded in {\sf EPR}.
Furthermore, it is also reasonable to expect the guarantee ($G$) of a
transaction to be expressible in the same logic as its inferred $\F$,
since $\F$ (without the stability check) is essentially a complete
characterization of the transaction, while $G$ is only an abstraction.
Thus, with $\SL[{\sf SLA},{\sf EPR}+{\sf SLA}]$ as the language of
inference, the verification problem for weakly isolated transactions
is decidable. Moreover, off-the-shelf SMT solvers (e.g., Z3) are
equipped with efficient decision procedures for ${\sf EPR}+{\sf SLA}$,
making automated verification a practical exercise.

>>>>

The rely-guarantee framework presented in the previous section
facilitates modular proofs for weakly-isolated transactions, but
imposes a non-trivial annotation burden.  In particular, it requires
each statement ($c$) of the transaction to be annotated with a stable
pre- ($P$) and post-condition ($Q$), and loops to be annotated with
stable inductive invariants ($\psi$). While weakest pre-condition
style predicate transformers can help in inferring intermediate
assertions for regular statements, loop invariant inference remains
challenging, even for the simple form of loops considered here.  As an
alternative, we present an inference algorithm based on state
transformers.  The idea is to infer the logical effect that each
statement has on the transaction-local database state $\stl$ (i.e.,
how it transforms $\stl$), and compose multiple such effects together
to describe the effect of the transaction as a whole.  Importantly,
this approach generalizes to loops, where the effect of a loop can be
inferred as a well-defined function of the effect of its body, thanks
to certain pleasant properties enjoyed by the database programs
modeled by $\txnimp$.  Interpreting database semantics as functional
transformations on sets (described in terms of their logical effects)
enables an inference mechanism that can leverage off-the-shelf SMT
solvers for automated verification.

@@

At the core of our approach is a simple language ($\SL$) to express
set transformations (see Fig.~\ref{fig:logic-syntax}). The language
admits only set expressions that include variables ($x$), literals of
the form $\{x \,|\, \varphi\}$ where $\varphi$ is a propositional
(quantifier-free) formula on $x$, a restricted form of existential
quantification that binds a set $x$ satisfying proposition $\phi$ in a
set expression $s$, a monadic composition of two set expressions
($s_1$ and $s_2$) composed using a bind ($\bind$) operation, a
conditional set expression where the condition is a propositional
formula, and a union of two set expressions. Symbols $\stl$ and $\stg$
are also variables in $\SL$, but are used to denote local and database
states, respectively.  The language is carefully chosen to be
expressive enough to capture the semantics of $\txnimp$ statements (as
well as SQL operations more generally), yet simple enough to have a
semantics-preserving translation to a decidable fragment of
first-order logic.

@@

where $\bar{v}$ are the variables that occur free in $\F$; this is

@@

equivalence in $\SL$ translates to equality in first-order logic,

@@

non-negatives ($I(\stg) \Leftrightarrow \forall(r\in\stg).~r.\C{bal}\ge

@@

The state transformer rules, like the earlier RG rules, closely follow
the corresponding reduction rules in Fig.~\ref{fig:txnimp}, except
that their language of expression is $\SL$. For instance, while the
reduction rule for \C{UPDATE} declaratively specifies the set of
updated records, the state transformer rule uses $\SL$'s bind
operation to \emph{compute} the set. Other SQL rules do likewise. The
rules for \C{LET} binders, conditionals, and sequences compose the
effects inferred for their subcommands. Thus, the effect of a sequence
of commands $c_1;c_2$ is the union of effects $F_1$ and $F_2$ of $c_1$
and $c_2$, respectively, except that $F_2$ is applied to a state
($\stl$) to which $F_1$ has already been applied, reflecting their
order of reduction. The inference rule for \C{FOREACH} takes advantage
of the $\SL$'s bind operator to lift the effect inferred for the loop
body to the level of the loop. Since records added to $\stl$ in each
iteration of \C{FOREACH} are independent of the previous iteration,
sequential composition of the effects of different iterations is same
as their parallel composition. Since the loop body is executed once
per each $z\in x$, the effect of the the loop is the union of effects
($\F$) for all $z\in x$, all applied to the same state ($\stl$ and
$\stg$). That is, $\F_{loop}(\stl,\stg) = \bigcup_{z\in
  x}\F_{body}(\stl,\stg)$. From the definition of the set monad's bind
operator, $\F_{loop}(\stl,\stg) = x \bind (\lambda
z.~F_{body}(\stl,\stg))$, which mirrors the definition of the rule.
This development mirrors ideas found in type checkers for certain
dependent type systems~\cite{KJ14}.

@@

Theorem~\ref{thm:inference-sound} lets us replace the local judgment
of the \rulelabel{RG-Txn} rule (Fig.~\ref{fig:rg-rules}) by a state
transformer inference judgment. The soundness of a transaction's guarantee can now
be established w.r.t the effect $\F$ of the body. The
\rulelabel{RG-Txn} rule so updated is shown
below:
\begin{smathpar}
\begin{array}{c}
\RULE
{
  \stable(R,\I)\spc
  \stable(R,I)\\
   \R_e(\stl,\stg,\stg') \Leftrightarrow \exists \stg_1. \I_e(\stl, \stg_1, \stg) \wedge R(\stg, \stg') \wedge \I_e(\stl, \stg_1, \stg') \spc  c \elabsto \F\\
  \R_c(\stl,\stg,\stg') \Leftrightarrow \exists \stg_1. \I_c(\stl, \stg_1, \stg) \wedge R(\stg, \stg') \wedge \I_c(\stl, \stg_1, \stg') \spc \stable(\R_c,\F)\\
  \forall \stg.~ G(\stg, F(\emptyset,\stg) \gg \stg)\spc
  \forall \stg,\stg'.~I(\stg) \wedge G(\stg,\stg') \Rightarrow I(\stg')\\
}
{
  \rg{I,R}{\ctxn{i}{\I}{c}}{G,I}
}
\end{array}
\end{smathpar}
Automating the application of the \rulelabel{RG-Txn} rule for a
transaction requires automating the multiple implication checks in
the premise. While $R$, $\R$, $\R_c$, $\I$ and $I$ are formulas in
first-order logic (FOL) with a relatively simple structure, $\F$
is an expression in the set language $\SL$
(Fig.~\ref{fig:logic-syntax}) with a possibly complex structure.
Fortunately, however, there exists a semantics-preserving translation
from $\SL$ to a restricted subset of first-order logic (FOL) that
lends itself to automatic reasoning.

@@

Observe that the encoding shown in Fig.~\ref{fig:logic} maps to
a subset of logic that satisfies the following syntactic properties:
\begin{itemize}
  \item All quantification is first-order; second-order objects, such
    as sets and functions, are never quantified.
  \item Quantifiers appear only at the prenex position, i.e., at the
    beginning of a quantified formula.  
  \item All function symbols, modulo those that might appear in
    $\varphi\in\Prop^0$, are uninterpreted and Boolean.
\end{itemize}
This fragment of FOL known as EPR (Effectively Propositional logic)
is known to be decidable~\cite{z3epr}. The language of encoding, however, is a
combination of {\sf EPR} with (a). $\Prop^0$, the theory from which
quantifier-free propositions ($\varphi$) that encode object language
expressions are drawn, and (b).  $\Prop^1$, the theory from which
invariants ($I$) are drawn. We write $\SL[\Prop^0,\Prop^1]$ to
highlight the parameterization of $\SL$ on $\Prop^0$ and $\Prop^1$,
and state the following theorem:
\begin{theorem}
  $\SL[\Prop^0,\Prop^1]$ is decidable if ${\sf EPR}+\Prop^0+\Prop^1$
  is decidable.
\end{theorem}
A useful instantiation of $\SL$ is $\SL[{\sf SLA},{\sf EPR}+{\sf
SLA}]$, where ${\sf SLA}$ is the theory of simple linear arithmetic.
Since {\sf EPR}+{\sf SLA} is known to be decidable~\cite{eprsla}:
\begin{theorem}
  $\SL[{\sf SLA},{\sf EPR}+{\sf SLA}]$ is decidable.
\end{theorem}
The $\SL[{\sf SLA},{\sf EPR}+{\sf SLA}]$ instantiation requires $I$ to
be drawn from {\sf EPR}+{\sf SLA}, which is expressive enough to
describe common database integrity constraints, such as referential
integrity, non-negativeness of all integer values in a column etc.
The isolation specifications presented in Sec.~\ref{sec:isolation} are
already simple first-order formulas that can be encoded in {\sf EPR}.
Furthermore, it is also reasonable to expect the guarantee ($G$) of a
transaction to be expressible in the same logic as its inferred $\F$,
since $\F$ (without the stability check) is essentially a complete
characterization of the transaction, while $G$ is only an abstraction.
Thus, with $\SL[{\sf SLA},{\sf EPR}+{\sf SLA}]$ as the language of
inference, the verification problem for weakly isolated transactions
is decidable. Moreover, off-the-shelf SMT solvers (e.g., Z3) are
equipped with efficient decision procedures for ${\sf EPR}+{\sf SLA}$,
making automated verification a practical exercise.

### 2017-07-08 reasoning.tex

\rulelabel{RG-Txn} is the top-level rule that lets us prove a
transaction preserves the high-level invariant $I$ when executed under
the required isolation as specified by $\I$. It relies on the
transaction-local judgment to verify the body ($c$) of a transaction
$i$. The precondition $P$ of $c$ must follow from the fact that the
transaction-local database ($\stl$) is initially empty, and the global
database satisfies the high-level invariant $I$. The rely relation
($\R_l$) for the local judgment is a ternary relation computed as $R$
modulo $\I_e$:
\begin{smathpar}
\begin{array}{lcl}
%\R_l(\stl,\stg,\stg') & \Leftrightarrow & R(\stg,\stg') \conj
%\I_e(\stl,\stg,\stg')
  \R_l(\stl,\stg,\stg') & \Leftrightarrow & \exists \stg_1. \I_e(\stl, \stg_1, \stg) \wedge R(\stg, \stg') \wedge \I_e(\stl, \stg_1, \stg')
\end{array}
\end{smathpar}
In words, if there exists $\stg_1$ which was the global database state
during the last step by a transaction and if the transaction can take
its next step at the current global database state $\stg$, then even
after the interference, the transaction should still be able to take
its next step. If there is no such state $\stg_1$, then the
transaction will be stuck, and the interference can be safely
discarded. Often, this significantly simplifies the form of $\R_l$
irrespective of $R$. For example, when a transaction is executed
against a snapshot of the database (i.e. $\mathbb{I}_{ss}$), $\R_l$
will be the identity function, since if $\stg \neq \stg'$, there does
not exist any $\stg_1$ which satisfies the constraint for
$\R_l(\stl,\stg,\stg')$.

@@

For the guarantee $G$ of a transaction to be valid, it must follow
from the post-condition $Q$ of the body, provided that $Q$ is stable
w.r.t the commit-time interference captured by $\R_c$. $\R_c$, like
$\R_l$, is computed as a rely relation modulo isolation, except that
commit-time isolation ($\I_c$) is considered. The validity of
$G$ is captured by the following implication:
\begin{smathpar}
\begin{array}{c}
  \forall \stl,\stg.~ Q(\stl,\stg) \Rightarrow G(\stg, \stl \gg \stg)\spc
\end{array}
\end{smathpar}
In other words, if $Q$ relates the transaction-local database state
($\stl$) to the state of the global database ($\stg$) before a transaction
commits, then $G$ must relate the states of the global database before
and after the commit. The act of commit is captured by the flush
operation ($\stl\gg\stg$). Once we establish the validity of $G$ as a
faithful representative of the transaction, we can verify that the
transaction preserves the high-level invariant $I$ by checking the
stability of $I$ w.r.t $G$, i.e., $\forall \stg,\stg'.~I(\stg) \wedge
G(\stg,\stg') \Rightarrow I(\stg')$.

>>>>

\rulelabel{RG-Txn} is the top-level rule that lets us prove a
transaction preserves the high-level invariant $I$ when executed under
the required isolation as specified by $\I$. It relies on the
transaction-local judgment to verify the body ($c$) of a transaction
$i$. The precondition $P$ of $c$ must follow from the fact that the
transaction-local database ($\stl$) is initially empty, and the global
database satisfies the high-level invariant $I$. The rely relation
($\R_e$) for the local judgment is a ternary relation computed as $R$
modulo $\I_e$:
\begin{smathpar}
\begin{array}{lcl}
%\R_e(\stl,\stg,\stg') & \Leftrightarrow & R(\stg,\stg') \conj
%\I_e(\stl,\stg,\stg')
  \R_e(\stl,\stg,\stg') & \Leftrightarrow & \exists \stg_1.  R(\stg, \stg') \wedge \I_e(\stl, \stg_1, \stg')
\end{array}
\end{smathpar}

While $R$ describes changes in the global database state, $\I_e$
constrain the global effects that will visible to the transaction.
Recall, however that $\I_e$ uses the global state in effect at the
point the transaction last took a step ($\stg_1$) to determine whether
the transaction can proceed from the current global state ($\stg$).
Moreover, the global rely relation $R$ defines the permissible
interferences from $\Delta$ to a future global state ($\Delta'$).  The
local rely relation $\R_e$ determines allowed interferences as
specified by $\I_e$ assuming $\stg_1$.  In effect, the existential on
$\stg_1$ defines this reley relation to be an over-approximation on
the interferences in the presence of which the transaction can
proceed, simplifying verification burden.

%Often, this significantly simplifies the form of $\R_l$ irrespective
%of $R$. For example, when a transaction is executed against a
%snapshot of the database (i.e. $\mathbb{I}_{ss}$), $\R_l$ will be the
%identity function, since if $\stg \neq \stg'$, there does not exist
%any $\stg_1$ which satisfies the constraint for
%$\R_l(\stl,\stg,\stg')$.

%In words, if there exists $\stg_1$ which was the global database state during the last step by a transaction and if the transaction can take its next step at the current global database state $\stg$, then even after the interference, the transaction should still be able to take its next step.

@@

For the guarantee $G$ of a transaction to be valid, it must follow
from the post-condition $Q$ of the body, provided that $Q$ is stable
w.r.t the commit-time interference captured by $\R_c$. $\R_c$, like
$\R_e$, is computed as a rely relation modulo isolation, except that
commit-time isolation ($\I_c$) is considered. The validity of
$G$ is captured by the following implication:
\begin{smathpar}
\begin{array}{c}
  \forall \stl,\stg.~ Q(\stl,\stg) \Rightarrow G(\stg, \stl \gg \stg)\spc
\end{array}
\end{smathpar}
In other words, if $Q$ relates the transaction-local database state
($\stl$) to the state of the global database ($\stg$) before a transaction
commits, then $G$ must relate the states of the global database before
and after the commit. The act of commit is captured by the flush
operation ($\stl\gg\stg$). Once we establish the validity of $G$ as a
faithful representative of the transaction, we can verify that the
transaction preserves the high-level invariant $I$ by checking the
stability of $I$ w.r.t $G$, i.e., $\forall \stg,\stg'.~I(\stg) \wedge
G(\stg,\stg') \Rightarrow I(\stg')$.

### 2017-07-08 related-work.tex

>>>>

\section{Conclusions}
\label{sec:conclusions}

To improve performance, modern database systems employ techniques that
weaken the strong isolation guarantees provided by serializable
transactions in favor of alternatives that allow a transaction to
witness the effects of other concurrently executing transactions that
happen commit during its execution.  Typically, it is the
responsibility of the database programmer to determine if an available
weak isolation level would violate a transaction's consistency
constraints.  Although this has proven to be a difficult and
error-prone process, there has heretofore been no attempt to formalize
notions of weak isolation with respect to application semantics, or
consider how we might verify the correctness of database programs that
use weakly-isolated transactions.  In this paper, we provide such a
formalization.  We develop a rely-guarantee proof framework cognizant
of weak isolation semantics, and build on this foundation to devise an
inference procedure that facilitates automated verification of
weakly-isolated transactions, and have applied our ideas on
widely-used database systems to justify their utility.  Our solution
enables database applications to leverage the performance advantages
offered by weak isolation, without compromising safety and
correctness.
